### 2019-05-27 treestats_paper.tex
examples

\newcommand{\nw}{{w}} % node weights
\newcommand{\iw}{{\tilde w}} % initial weights
\newcommand{\aw}{{\bar w}} % allele weights
\newcommand{\Tw}{{W}} % allele weights

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Theoretical framework}

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome.
\jk{Maybe use $n$ here? We use it below}
Suppose first that we wish to compute an additive function of the allele frequency spectrum.
In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
and that the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
This definition applies to multiallelic sites by assuming that for each allele
we can only see how many samples carry that allele,
not how the remaining samples are partitioned between one or more other alleles.
\jk{Not clear to me here if we're talking about frequency $0 \leq p \leq 1$ or
`prevalance`, $0 \leq p \leq N$.}
(Note also that this does not differentiate ancestral and derived alleles,
and so is a function of the \emph{folded} allele frequency spectrum.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and removing those that subsequently mutate).

\jk{Big jump here. I don't see the connection between the previous paragraph
and the bitstring representation of haplotype statistics}
There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
and so a general statistic of biallelic genotypes could be represented by a vector of length $2^n$.
A more convenient way to parameterize many functions of genotype patterns
uses three ingredients:
\begin{itemize}

@@

    \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that
        $F(\nw)$ is the ``summary'' statistic we compute \jk{I'd find it more
        intuitive to use $f$ as the notation for a summary function. We can use
        something else to define the site frequency spectrum above, right? Big
        $F$ says ``set" to me.}

@@

% Sum weights for each allele.
Concretely, to compute the statistic at a given site $j$,
for each allele $a$ seen at this site,
let $\aw_j(a)$, the weight associated with this allele,
be the sum of the initial weights over all nodes who inherit allele $a$.
If the site has only one mutation, which produced allele $a$
at node $i$, then $\aw_j(a) = \nw(i)$.
Then, the statistic computed at site $j$ is
\begin{align}
    \site(F, \iw)_j
    &=
    \sum_a \aw_j(a) ,
\end{align}
and a summary statistic across $L$ sites is
\begin{align}
    \bar \site(F, \iw)
    &=
    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
\end{align}
We call this the \textbf{site statistic} associated with $\iw$ and $F$.

@@

@@

\plr{Introduce branch stats.}
The \textbf{branch statistic} associated with the branch of length $t$
above node $n$ on a tree of length $\ell$,
with total weight $\Tw = \sum_j \iw(j)$
and node weights $\nw$,
is
\begin{align}
    \branch(F, \iw)_n
    &=
    \ell t ( F(\Tw - \nw(n)) + F(\nw(n)) ) .
\end{align}
This is the summary value that would be added to a site statistic
if a single mutation occurred on this branch:
we must apply the summary function
to both the weight below $n$ as well as the weight \emph{not} below $n$
to account for both derived and ancestral alleles.

\paragraph{Summary}
We have defined three types of statistic using the same weighting scheme:
\begin{enumerate}
    \item \textbf{Site statistics}
        -- summarizes weights below each allele at each polymorphic site.
    \item \textbf{Node statistics}
        -- summarizes total weight below each node.
    \item \textbf{Branch statistics}
        -- summarizes total weight below each branch, multiplied by branch length.
\end{enumerate}

@@

    \item How will polarization work?
        Right now, Site and Branch stats are polarized but Node statistics are not.
        (To make Node statistics unpolarized, we need to also add $F(\Tw - w)$
        to the weight of each node.)
        Perhaps this is OK -- Site and Branch stats should correspond to each other,
        but Node statistics are different, so could have different behavior.
        BUT, people will need to compute polarized Site and Branch stats;
        we need to make it easy for them somewhere.

    \item What about normalization?
        In the initial implementation, Node statistics were divided by the length
        for which the node was an ancestor to any sample, rather than just total sequence length
        To make these more analogous to the other stats, we think we want to remove this.
        The denominator for the original case is easily computable, by using
        $F(x) = 1$ if $x>0$ and $F(x)=0$ otherwise, with appropriate (positive) weights.

@@

Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
is the genotype of the $a^\text{th}$ individual
at the $j^\text{th}$ site, with `1` denoting the derived state
(we assume here biallelic markers).
Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
Then the \emph{genetic covariance} between samples $a$ and $b$ is
\begin{align*}
    C_{ab}
        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
\end{align*}
This can be computed by letting
$\iw_a = \delta_a$ and $\iw_b = \delta_b$
(these weights mark nodes above $a$ and $b$ respectively)
and $\iw_t = \bone / n$
(this weight gives the proportion of samples below each node).
Then, with
$F(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
we can compute the covariance as
\begin{align*}
    C_{ab} = \frac{1}{2} \bar \site(F, (\iw_a, \iw_b, \iw_t)) .
\end{align*}
This follows, including the factor of two,
because the summary function applied to the ancestral allele
at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
in sample $a$, sample $b$, and total, respectively, is
$F(1 - p_a, 1 - p_b, 1 - p_t) = F(p_a, p_b, p_t)$.

Computing the full covariance matrix for a large number of samples may be infeasible.
However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
This works because if we set the initial weights to $y$,
then each allele's weight is equal to the sum of the entries of $y$
corresponding to samples carrying that allele.
Then, if we define
$F_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
then
\begin{align} \label{eqn:yCz}
    \frac{1}{2} \bar \site(F_{yz}, (y, z, \iw_t))
        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)
                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
        &= y^T C z .
\end{align}

>>>>

\newcommand{\nw}{x} % node weights
\newcommand{\iw}{w} % initial weights
\newcommand{\aw}{{\bar w}} % allele weights
\newcommand{\Tw}{{W}} % allele weights

@@

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome.
\jk{Maybe use $n$ here? We use it below}
Suppose first that we wish to compute an additive function of the allele frequency spectrum.
In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
and that the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
This definition applies to multiallelic sites by assuming that for each allele
we can only see how many samples carry that allele,
not how the remaining samples are partitioned between one or more other alleles.
\jk{Not clear to me here if we're talking about frequency $0 \leq p \leq 1$ or
`prevalance`, $0 \leq p \leq N$.}
(Note also that this does not differentiate ancestral and derived alleles,
and so is a function of the \emph{folded} allele frequency spectrum.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and removing those that subsequently mutate).
What we do next is a generalization of this idea.

A more convenient way to parameterize many functions of genotype patterns
uses three ingredients:
\plr{MODIFY and SHORTEN THIS}
\begin{itemize}

@@

    \item[(c)] A \emph{summary function} $f : \R^k \to \R$ so that
        $f(\nw)$ is the ``summary'' statistic we compute \jk

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Subtree weights}

We could calculate for a given tree how many of 
a certain set of ``target'' samples are below each of its edges
by (a) assigning each of those samples weight 1, and other nodes weight zero, then
(b) finding the total weight of all nodes in the subtree below each edge.
This total weight is equal to the number of target samples below that edge,
and it can be calculated efficiently by propagating weights up the tree.
It turns out to be useful to allow more general weights.

\begin{definition}[Subtree weights]
    Given a list of \emph{node weights} $w$, and a tree $T$,
    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all node weights
    of every node below $u$ in the tree (including $u$ itself):
    \begin{align*}
        \nw_u = \sum_{v \,:\, v \le_T u} \iw_v ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ back to the root of $T$.
\end{definition}

We can easily update the vector $x$ as we move along the genome
XXX as follows or described later.

Next, we will describe several ways to summarize these subtree weights.

%%%%%%%%%%%%%%%%%%
\subsection*{Site statistics}

The first method of summarizing uses subtree weights
to compute a summary for each variant site in the genome.
To compute the statistic at site $j$,
for each allele $a$ seen at this site,
let $\aw_j(a)$, the weight associated with this allele,
be the sum of the initial weights over all nodes who inherit allele $a$.
Note that if there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the subtree weight of the node that the mutation producing $a$
appeared above, 
while more generally $\aw$ is not necessarily a subtree weight 
but is easily computable from them.
If the site has only one mutation, which produced allele $a$
at node $i$, then $\aw_j(a) = \nw(i)$.
Then, the statistic computed at site $j$ is
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a \in D_j} \aw_j(a) ,
\end{align}
where the sum is over the \emph{derived} alleles $D_j$ at site $j$.
A summary statistic across $L$ sites is
\begin{align}
    \bar \site(f, \iw)
    &=
    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
\end{align}
We call this the \textbf{site statistic} associated with $\iw$ and $f$.

\paragraph{Why the derived alleles?}
We have defined the statistic to sum over only derived alleles,
i.e., alleles not shared with the root of the tree.
At a biallelic site, this is equivalent to knowing the derived allele frequency
among any group of samples,
and such a statistic will be unchanged under recoding
if the summary function is unchanged after exchanging $p$ for $1-p$.
(XXX what is $p$)

\begin{example}[Nucleotide diversity]
    Suppose we want to compute the mean density of nucleotide differences
    in a group $S$ of $n$ sequences.
    To do this, let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    so that $\nw_u$ gives the number of nodes in $S$ below $u$,
    and define
    \begin{align*}
        f(x) = \frac{2 x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\bar \site(f, \iw)$ is mean nucleotide diversity in the region.
\end{example}

\plr{note LATER about what this does with triallelic sites}

\begin{example}[Nucleotide divergence]
    Now Suppose we want to compute the mean density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before, let $\iw_{ju} = 1$ for $u \in S_j$ and $\iw_{jv} = 0$ otherwise,
    so that $\nw_{ju}$ gives the number of nodes in $S_j$ below $u$.
    Then we define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1 (n_2 - x_2) + (n_1 - x_1) x_2}{n_1 n_2} .
    \end{align*}
    Then $\bar \site(f, \iw)$ is mean nucleotide divergence between the two in the region.
\end{example}

\begin{example}[Phenotypic correlations]
    Suppose that each sample has a phenotypic value, denoted $p_u$ for $u \in S$,
    and we want to compute for each site the correlation of phenotype and genotype.
    Suppose we have normalized $p$ so that $\sum_u p_u = 0$ and $\sum_u p_u^2 = 1$.
    Then, if $g$ is a vector of binary genotypes (so $g_u = 1$ if $u$ carries the derived allele),
    then the covariance of $p$ with $g$ is just $\sum_u p_u g_u = \sum_{u \st g_u = 1} p_u$,
    i.e., the sum of the phenotypes of all samples carrying the derived allele.
    This means that if we define $\iw_{1u} = p_u$, and $\iw_{2u} = 1$ for $u \in S$ and zero otherwise,
    and $f(x_1, x_2) = x_1 / x_2$,
    then $\site(\iw, f)_j$ is the correlation between $p$ and the allele at site $j$.
\end{example}

\begin{example}[Patterson's $f_4$]
    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
    is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
    Therefore, if as before we let $\iw_{ju} = 1$ if $u \in S_j$,
    and $f(x_1, x_2, x_3, x_4) = (x_1 n_2 - x_2 n_1) (x_3 n_4 - x_4 n_3) / n_1 n_2 n_3 n_4$,
    then $\bar \site(\iw, f)$ gives Patterson's $f_4$ across the region.
\end{example}

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Branch statistics}

\plr{Introduce branch stats.}

The \textbf{branch statistic} associated with the branch of length $t$
above node $u$ on a tree of length $\ell$,
with subtree weights $x$,
is
\begin{align}
    \branch(f, \iw)_u
    &=
    \ell t  f(\nw(u))  .
\end{align}
This is the summary value that would be added to a site statistic
if a single mutation occurred on this branch,
multiplied by the expected number of mutations 
(up to an overall constant related to mutation rate).

\begin{example}[Mean TMRCA]
    If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
    then $f(u)$ gives the probability that the branch above $u$
    lies on the path from two randomly chosen samples from $S$
    on the path up to their most recent common ancestor (MRCA).
    Therefore, $\bar \branch(f, \iw)$ for these choices
    gives the mean total distance in the tree between two samples from $S$,
    averaged across the sequence.
    (This is twice the TMRCA if the samples are all from the same time.)
\end{example}

\begin{example}[Phenotypic correlation with pedigree]
    If we take $\iw$ and $f$ as in the ``Phenotypic correlations'' example above,
    what does $\branch(f, \iw)$ tell us?
    The statistic tells us the \emph{expected} correlation between phenotype and any mutations
    appearing on the tree, which is a summary of who much local relatedness
    aligns with similarity in phenotype.
\end{example}

\begin{example}[Patterson's $f_4$]
    Suppose that the four subsets each consist of only a single sample.
    The summary function $f(x_1, x_2, x_3, x_4)$ for the $f_4$ statistic
    then assigns weight 1 to any branch that separates $x_1$ and $x_3$ from $x_2$ and $x_4$,
    and weight -1 to any branch that separates $x_1$ and $x_4$ from $x_2$ and $x_3$.
    The statistic $\branch(f, \iw)$ therefore 
    gives the difference in averages of these two types of branch,
    averaged over choices of samples from the sample sets and averaged across the genome.
\end{example}

%%%%%%%%%%%%%%%%%%
\subsection*{Node statistics}

\textbf{Suppose we output per node.}

@@

\begin{example}[Ancestry proportions]
    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
    and then $\nw_u / n$ is the proportion of the samples in $S$ that lie below $u$.
    Therefore, if $f(x) = x / n$,
    then $\node(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Summary}
We have defined three types of statistic using the same weighting scheme:
\begin{enumerate}
    \item \textbf{Site statistics}
        -- summarizes weights below each allele at each polymorphic site.
    \item \textbf{Node statistics}
        -- summarizes total weight below each node.
    \item \textbf{Branch statistics}
        -- summarizes total weight below each branch, multiplied by branch length.
\end{enumerate}

@@

    \item What about normalization?
        In the initial implementation, Node statistics were divided by the length
        for which the node was an ancestor to any sample, rather than just total sequence length
        To make these more analogous to the other stats, we think we want to remove this.
        The denominator for the original case is easily computable, by using
        $f(x) = 1$ if $x>0$ and $f(x)=0$ otherwise, with appropriate (positive) weights.

@@

Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
is the genotype of the $a^\text{th}$ individual
at the $j^\text{th}$ site, with `1` denoting the derived state
(we assume here biallelic markers).
Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
Then the \emph{genetic covariance} between samples $a$ and $b$ is
\begin{align*}
    C_{ab}
        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
\end{align*}
This can be computed by letting
$\iw_a = \delta_a$ and $\iw_b = \delta_b$
(these weights mark nodes above $a$ and $b$ respectively)
and $\iw_t = \bone / n$
(this weight gives the proportion of samples below each node).
Then, with
$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
we can compute the covariance as
\begin{align*}
    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
\end{align*}
This follows, including the factor of two,
because the summary function applied to the ancestral allele
at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
in sample $a$, sample $b$, and total, respectively, is
$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.

Computing the full covariance matrix for a large number of samples may be infeasible.
However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
This works because if we set the initial weights to $y$,
then each allele's weight is equal to the sum of the entries of $y$
corresponding to samples carrying that allele.
Then, if we define
$f_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
then
\begin{align} \label{eqn:yCz}
    \frac{1}{2} \bar \site(f_{yz}, (y, z, \iw_t))
        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)
                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
        &= y^T C z .
\end{align}
